# Introduction

The `rsed` package contains tools for stream editing:  manipulating text files by making insertions, replacements, deletions, substitutions, or commenting. It hails from the powerful Unix command, `sed`.  See, for example, the [GNU documentation for `sed`](https://www.gnu.org/software/sed/manual/sed.html).  While this package is not nearly as powerful as `sed`, it is much simpler to use. Just because `rsed` was developed to mimic some of the functionality of a Unix command, be advised it works on all the standard platforms, including Windows.

I find `rsed` to be particularly useful in workflows that create text files (e.g. html or otherwise) that need to be tweaked a little before they can be used. For example, I use it when files of interest are produced by other packages, and I don't want to delve into the (complicated) source code in order to produce a slightly different result in the output text files.  In this vignette I'll take you through a simple tour of the capability of the `rsed` package.  Additional details can be found in the package [help files](rd.html).

## What's new here?

You might ask, "doesn't this package duplicate functionality already available in R?  Can't I perform these same tasks using base R functions like [readLines()](http://www.inside-r.org/r-doc/base/readLines), [gsub()](http://www.inside-r.org/r-doc/base/gsub), [grep()](http://www.inside-r.org/r-doc/base/grep), etc.?"  The answer is yes.  That being said, the value of the `rsed` package is that it collapses the several steps of code that would be required for stream editing into a call to a single function.

## Package installation

Package installation is made easy with [devtools](https://cran.r-project.org/web/packages/devtools/index.html):
```{r instal, eval = FALSE, echo = TRUE}
# If devtools is not already installed
install.packages("devtools") 

devtools::install_github("pnnl/rsed")
```
Now load the package as usual:
```{r loadpackage, eval = TRUE, echo = TRUE}
library(rsed)
```
## What's a stream?

For our purposes, a *stream* is simply a character vector. The mental model is that each element of the vector corresponds to a line in a file.
While this vignette and the package documentation typically discuss stream editing in terms of manipulating files stored on disk, reads and writes to disk are not required.  The functions in `rsed` will operate on any stream (i.e. any character vector).  

Here's an example of a stream, from *To a Skylark*, by Percy Bysshe Shelley:

```{r stream, eval = TRUE, echo = TRUE}
skylark <- c("Higher still and higher",
             "From the earth thou springest",
             "Like a cloud of fire;",
             "The blue deep thou wingest,",
             "And singing still dost soar, and soaring ever singest.")

# To view it as a stream:
as.stream(skylark)
```
All of the functions in `rsed` operate on and return character vectors like `skylark`.

# Stream editing functions

`rsed` can perform five types of editing:

- Insertion of new lines at any point in the file
- Replacement of entire lines
- Deletion of entire lines
- Substituting one string for another
- Comment out a line using the syntax of any programming language

We'll look at each in turn.  To avoid marring Shelley's lovely verse, let's create something we'd like to edit and write it to a file:

```{r stream1, eval = TRUE, echo = TRUE}
cat("Add something after this line\n",
    "Replace this line\n",
    "Delete this line\n",
    "Substitute 'this' for 'that'\n",
    "Comment this line\n",
    file = "stream.txt", sep = "")
```
The file `stream.txt` looks like this:
```{r showStream, eval = TRUE, echo = FALSE}
cat("Add something after this line",
    "\nReplace this line",
    "\nDelete this line",
    "\nSubstitute 'this' for 'that'",
    "\nComment this line")
```
To illustrate how processing this stream would work if it began as a file, we would need to read the file and produce a *stream* using `readLines()`, such that each line of the file is an element in a character vector:

```{r stream2, eval = TRUE, echo = TRUE}
s <- as.stream(readLines("stream.txt"))
```

Now we can operate on the stream `s`.

## Insertion
Inserting new lines is accomplished by `r rdl("sed_insert()")`.  So, for easy comparison, here's the original stream, `s`:
```{r orig1, eval = TRUE, echo = FALSE}
s
```
And here's what we get after the insertion:
```{r insert, eval = TRUE, echo = TRUE}
sed_insert(s, "Add something after this line", "Here's the addition")
```
## Replacement
Replacing an entire line is accomplished by `r rdl("sed_replace()")`. Again for easy comparison, here's the original stream, `s`:
```{r orig2, eval = TRUE, echo = FALSE}
s
```
And after the replacement:
```{r replace, eval = TRUE, echo = TRUE}
sed_replace(s, "Replace this", "A replacement")
```

## Deletion
Deleting an entire line is also accomplished by `r rdl("sed_replace()")`.  The original stream, `s`:
```{r orig3, eval = TRUE, echo = FALSE}
s
```
and after the deletion:
```{r delete, eval = TRUE, echo = TRUE}
sed_replace(s, "Delete this", NULL)
```

## Substitution
Substitution is searching for a string and replacing it with another.  We can do this with `r rdl("sed_substitute()")`.  The original stream `s`:
```{r orig4, eval = TRUE, echo = FALSE}
s
```
After substitution:
```{r substitute, eval = TRUE, echo = TRUE}
sed_substitute(s, "'that'", "'this'")
```

## Commenting
We can use `r rdl("sed_comment()")` to add or remove comment characters from a number of languages to one or more selected lines. The supported languages are R, C, Java, html, tex, and SAS.  However, any commenting characters can be provided.  The original stream, `s`:
```{r orig5, eval = TRUE, echo = FALSE}
s
```
And after commenting the last line:
```{r comment, eval = TRUE, echo = TRUE}
sed_comment(s, "Comment", type = "html")
```

# Hi-level wrapper

All the edits shown in the previous section could be accomplished using the following steps:
```{r manual, eval = TRUE, echo = TRUE}
s <- readLines("stream.txt")
s <- sed_insert(s, after = "Add something", insertion = "Here's the addition")
s <- sed_replace(s, at = 3, replacement = "A replacement")
s <- sed_replace(s, at = 4, replacement = NULL)
s <- sed_substitute(s, pattern = "'that'", replacement = "'this'")
s <- sed_comment(s, at = "Comment", type = "html")
writeLines(s, con = "newStream1.txt")
```
Alternatively, we could use the hi-level function, `r rdl("streamEdit()")`, to execute all of these steps with a single function call:
```{r streamEdit, eval = TRUE, echo = TRUE}
streamEdit(list(i = list(after = "Add something", insertion = "Here's the addition"),
                r = list(at = 3, replacement = "A replacement"),
                r = list(at = 4, replacement = NULL),
                s = list(pattern = "'that'", replacement = "'this'"),
                c = list(at = "Comment", type = "html")),
            inFile = "stream.txt", 
            outFile = "newStream2.txt")
```
So here's the original:
```{r orig6, eval = TRUE, echo = TRUE}
as.stream(readLines("stream.txt"))
```
And here are the changes:
```{r big, eval = TRUE, echo = TRUE}
as.stream(readLines("newStream2.txt"))
```
And for good measure, let's compare the results of these two approaches:
```{r compare, eval = TRUE, echo = TRUE}
# The result from calling each function individually
s1 <- readLines("newStream1.txt")

# The result from the single call to streamEdit()
s2 <- readLines("newStream2.txt")

identical(s1, s2)
```

# The 'stream' class and 'print' method

`r rdl("as.stream()")` converts a character vector to a `stream` object by adding `stream` to it's class:
```{r streamClass, eval = TRUE, echo = TRUE}
s <- readLines("stream.txt")
class(s)
class(as.stream(s))
```
The only method for the `stream` class is `print`, which makes reading the stream easier. Consider printing without the `stream` class:
```{r print1, eval = TRUE, echo = TRUE}
print(s)
```
versus printing an object with the `stream` class:
```{r print2, eval = TRUE, echo = TRUE}
print(as.stream(s))
```
Note that character vectors do not need to be converted to streams via `r rdl("as.stream()")` before passing them to the stream editing functions.  Either way works:
```{r compare2, eval = TRUE, echo = TRUE}
sed_comment(s, 5, type = "R")
sed_comment(as.stream(s), 5, type = "R")
```

```{r unlink, eval = TRUE, echo = FALSE}
unlink(c("stream.txt", "newStream1.txt", "newStream2.txt"))
```

# A relevant example
The online documentation of the `rsed` package is produced by [packagedocs](https://github.com/hafen/packagedocs) (thank you Ryan Hafen!).  For some reason however, on the `r rdl("as.stream()")` help page, the usage for the `print` method is incorrectly rendered like this:

    printstream(x, ...)

A quick fix to this problem was to essentially add the following to the end of the documentation [build script](https://github.com/pnnl/docs-rsed/blob/gh-pages/build.R):

    streamEdit(list(s = list(pattern = "printstream(", 
                             replacement = "print(", 
                             fixed = TRUE)),
               inFile = "rd.html", outFile = "rd.html")


# Citation

Please cite the `rsed` package using the following reference:

Sego LH. 2016. rsed: Stream editing in R. Pacific Northwest National Laboratory. https://pnnl.github.io/rsed.
